PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-06-07 draft.tex
Responses to Richard's comments plus a few small changes

\ek{With regard to the `SP' vs. Schr{\"o}dinger-Poisson issue: I also don't like too many initialisms in publications but felt that in this case, due to the sheer number of instances of the term, and given that it seems to have been adopted as a standard by some of the other ULDM folk (e.g. arXiv 1801.03507, 1608.00802, 1801.04864), it made sense to compactify things. I've removed all instances of `SP' in this version, but feel that somehow the full term can appear rather cumbersome at times. What do you think?}
\re{Fair enough, but I think it reads better without -- ULDM works via analogy with CDM; it's not consistent but seems better to me. Also, have added and "EK" comment :-) }

@@

The existence of an extremely light scalar field, minimally coupled to gravity, is the central premise on which ULDM models are predicated. Within the ULDM framework, it is proposed that cosmological structure formation proceeds through the Bose-Einstein condensation of this scalar field. In this Section  \re{I would normally give Section a capital letter in this context} \ek{I would be inclined to only capitalise `section' in instances where it appears as a proper noun, i.e. "In Section X we showed...", where we are referring to a specifically named section, numbered X. It seems to me that in this particular context `section' is a common noun. Happy to capitalise all instances of the word if this is your recommendation.} 
\re{I agree, but thought ``this Section'' was an example of it being used as a proper noun. But it is not a huge deal either way.}
we derive the (non-relativistic) dynamical equations of the single macroscopic condensate wavefunction, namely, the Schr{\"o}dinger-Poisson equations. We begin with the action functional for a scalar field, $\phi$, minimally coupled to gravity and in the absence of self-interactions,
%
\begin{equation}\label{eq:action}
    S=\int \frac{d^4x}{\hbar}\sqrt{-g}\bigg\{\frac{1}{2}g^{\mu\nu}\partial_\mu\phi\partial_\nu\phi-\frac{m^2}{2\hbar^2}\phi^2\bigg\},
\end{equation}
%
where we have taken $c=1$ but retain factors of $\hbar$ at this stage. Applying the variational principle to this action yields the Euler-Lagrange equations
%
\begin{equation}\label{eq:e-l}
    \frac{1}{\sqrt{-g}}\partial_\mu\big[\sqrt{-g}g^{\mu\nu}\partial_\nu\phi\big]-\frac{m^2}{\hbar^2}\phi=0.
\end{equation}
We evaluate equation \ref{eq:e-l} using linear perturbation theory, adopting the perturbed FRW metric in the Newtonian gauge:
\begin{equation}\label{eq:pFRW}
    ds^2=-\big(1+2\Phi(\vec{r})\big)dt^2+a(t)^2\big(1-2\Phi(\vec{r})\big)d\vec{r}^{\,2}.
\end{equation}
To linear order in $\Phi(\vec{r})$ we obtain 
\begin{equation}\label{eq:linear}
    \Ddot{\phi}-\frac{\big(1+4\Phi(\vec{r})\big)}{a(t)^2}\nabla^2\phi+3\big(1+2\Phi(\vec{r})\big)H\Dot{\phi}+\big(1+2\Phi(\vec{r})\big)\frac{m^2}{\hbar^2}\phi=0,
\end{equation}
where $H=\Dot{a}(t)/a(t)$. At late times in an expanding universe, $H\ll m$ and it is sufficient to set $H=0$ and $a(t)=1$ in equation \ref{eq:linear}. This is a good approximation even at relatively high redshifts, including the epochs of early structure formation. Alternatively, if we consider a non-expanding universe, these equalities are true by definition. In either case, we can remove the second term in equation \ref{eq:linear} entirely, such that all remaining derivatives are second order in space or time.

In the non-relativistic limit, Equation \ref{eq:linear} can be treated using the WKB approximation\footnote{For a detailed explication of the WKB approximation in the non-relativistic limit, see \cite{Young2015}.}. This allows us to write an ansatz solution for the field $\phi$:
\begin{equation}\label{eq:ansatz}
    \phi=\frac{\hbar}{\sqrt{2}m}\big(\psi e^{-imt/\hbar}+\psi^* e^{imt/\hbar}\big),
\end{equation}
where $\psi$ is assumed to be slowly varying in the sense that $m\vert\psi\vert\gg\hbar\vert\dot{\psi}\vert$, $m\vert\dot{\psi}\vert\gg\hbar\vert\Ddot{\psi}\vert$, $m\vert\psi\vert\gg\hbar\vert\nabla{\psi}\vert$, and $m\vert\nabla{\psi}\vert\gg\hbar\vert\nabla^2{\psi}\vert$. Direct substitution of the ansatz solution into equation \ref{eq:linear}, discarding the heavily suppressed second derivative terms, yields
\begin{equation}\label{eq:schrodinger}
    i\hbar\Dot{\psi}=-\frac{\hbar^2}{2m}\nabla^2\psi+m\Phi\psi.
\end{equation}
We have thus shown that $\psi$ satisfies the Schr{\"o}dinger equation in this limit, which is interpreted as the unique macroscopic wavefunction of a Bose-Einstein condensate. It follows that the particle density of the condensate is given by $\vert\psi\vert^2$, so its mass density is simply $m\vert\psi\vert^2$. The local gravitational potential thus satisfies the Poisson equation,
\begin{equation}\label{eq:poisson}
    \nabla^2\Phi=4\pi G m \vert\psi\vert^2,
\end{equation}
where $G$ is Newton's gravitational constant. The coupled equations \ref{eq:schrodinger} and \ref{eq:poisson} together form the nonlinear Schr{\"o}dinger-Poisson system which describes the dynamics of ULDM in the non-relativistic regime. It is this system of equations which \PyUltraLight is designed to solve.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth, trim={0 0 0 0},clip]{phase_comparison}
  \caption{Head-on collisions between two solitons of mass 5, relative velocity 6, and initial separation 4 in code units. Time progresses from left to right across each row. \re{maybe a few more words on what is going on (e.g. we see that in the lower panel the solitons slow as they approach one another. Also, should we choose masses/separations where it does not take  20+ Gyr for them to approach one another. As a test of the dynamics it is not a big deal, but it looks a bit funny if it is several times the age of the universe}}
  \label{fig:repulsion}
\end{figure}

As demonstrated in \cite{Paredes2016}, the wavelike properties of ULDM can give rise to effective forces which  dramatically affect the dynamics of core collisions. These effective forces arise as a result of interference phenomena, rather than because of any local interactions the ULDM model might  incorporate.  Figure \ref{fig:repulsion} shows the results of two \PyUltraLight where two solitons of equal mass undergo head-on collisions. Contours of the density profile along the plane of symmetry are displayed. In one case (top) there is no phase offset between the initial solitons, while in the second  the phases differ by $\pi$. In the latter case, the phase shift creates an effective repulsive force between the two solitons.  Further discussion of this phenomenon and its possible observational consequences can be found in \cite{Paredes2016}.

@@

An extensive study of the tidal disruption of ULDM solitonic cores orbiting a central potential has recently been undertaken in \cite{Du2018} and we reproduce just one of their results here. When the Schr{\"o}dinger-Poisson system is recast as a system of hydrodynamical equations, the fluid velocity  at any point is the gradient of the smoothly-varying phase.  Since the curl of a gradient vanishes identically, the velocity field of the Schr{\"o}dinger-Poisson system must be irrotational, and $\nabla\times\vec{v}=0$. If a radially symmetric soliton is initialised in a circular orbit around a Newtonian potential, there will be an initial transient as the spherical profile becomes elongated along the radial direction of the central potential. Meanwhile, the velocity field corresponding to the overall orbital motion of the soliton is superposed with the internal velocity field, combining so as to produce a net flow with vanishing curl.

The family of Riemann-S ellipsoids describe uniformly rotating bodies which satisfy the condition of irrotationality \cite{Chandrasekhar1965}, and it is therefore the characteristic internal velocity field of such an ellipsoid which we expect to arise during the simulation. It is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann-S ellipsoid when orbiting a host mass. We a reproduce this result using \PyUltraLight, with Figure \ref{fig:riemann} showing the internal velocity field of a solitonic satellite after one complete revolution around a host mass. The soliton has indeed become elongated along the radial line connecting it to the host, indicating that it is tidally locked, while the internal velocity field can be seen to be that of an irrotational Riemann-S ellipsoid \cite{Daller2012}.\re{Did you actually check numerically it is a Riemann-S ellipsoid or is that a qualitative assessment? No worries; just wanting to be clear -- as I said previous, I am not sure most people know intuitively what a Riemann-S ellipsoid is. Or maybe I am just poorly educated.}

@@

@@

Figures \ref{fig:combined_1} and \ref{fig:binary} demonstrate energy conservation for two scenarios. The first case  shows the evolution of the energy of a single soliton undergoing significant tidal disruption within a Newtonian central potential. For this simulation a soliton of mass 12 in code units was initialised at a radial distance of 3 code units from the centre of a Newtonian central potential generated by a central mass of 1000 code units. As the soliton is disrupted, the kinetic energy increases, while the gravitational energy due to the central potential decreases, as expected. Meanwhile, the gravitational potential energy from self-interactions gradually increases toward zero as the disruption continues and the ULDM is halo spread over a greater area. In this case the sum of the individual energy components is conserved to $10^{-2}\ \%$  at a resolution of $256^3$. Higher grid resolutions result in even better energy conservation.

@@

@@

The current implementation of \PyUltraLight is already a useful tool for simulating dynamical ULDM systems and exploring their dynamics However, there is much scope for  improvement. In particular, future releases may incorporate a variable timestep and more sophisticated physics, including explicit self-interactions in the axion sector or additional matter components.  \PyUltraLight is publicly available under a BSD license.  \re{took some stuff out; no  need to tip our hand completely.}

@@

\re{Check the bib options so we don't see all the Planck authors :-) }

>>>>

@@

The existence of an extremely light scalar field, minimally coupled to gravity, is the central premise on which ULDM models are predicated. Within the ULDM framework, it is proposed that cosmological structure formation proceeds through the Bose-Einstein condensation of this scalar field. In this Section  \re{I would normally give Section a capital letter in this context} \ek{I would be inclined to only capitalise `section' in instances where it appears as a proper noun, i.e. "In Section X we showed...", where we are referring to a specifically named section, numbered X. It seems to me that in this particular context `section' is a common noun. Happy to capitalise all instances of the word if this is your recommendation.} 
\re{I agree, but thought ``this Section'' was an example of it being used as a proper noun. But it is not a huge deal either way.}\ek{The way I would tend to think of it is by analogy with the following: "This person studies physics." vs. "Emily studies physics." In the former sentence, "person" is used as a description rather than a title, so would not be capitalised. Similar to "Richard is a professor." vs. "This is Professor Richard."}
we derive the (non-relativistic) dynamical equations of the single macroscopic condensate wavefunction, namely, the Schr{\"o}dinger-Poisson equations. We begin with the action functional for a scalar field, $\phi$, minimally coupled to gravity and in the absence of self-interactions,
%
\begin{equation}\label{eq:action}
    S=\int \frac{d^4x}{\hbar}\sqrt{-g}\bigg\{\frac{1}{2}g^{\mu\nu}\partial_\mu\phi\partial_\nu\phi-\frac{m^2}{2\hbar^2}\phi^2\bigg\},
\end{equation}
%
where we have taken $c=1$ but retain factors of $\hbar$ at this stage. Applying the variational principle to this action yields the Euler-Lagrange equations
%
\begin{equation}\label{eq:e-l}
    \frac{1}{\sqrt{-g}}\partial_\mu\big[\sqrt{-g}g^{\mu\nu}\partial_\nu\phi\big]-\frac{m^2}{\hbar^2}\phi=0.
\end{equation}
We evaluate equation \ref{eq:e-l} using linear perturbation theory, adopting the perturbed FRW metric in the Newtonian gauge:
\begin{equation}\label{eq:pFRW}
    ds^2=-\big(1+2\Phi(\vec{r},t)\big)dt^2+a(t)^2\big(1-2\Phi(\vec{r},t)\big)d\vec{r}^{\,2}.
\end{equation}
To linear order in $\Phi(\vec{r})$ we obtain 
\begin{equation}\label{eq:linear}
    \Ddot{\phi}-\frac{\big(1+4\Phi\big)}{a(t)^2}\nabla^2\phi+3H\Dot{\phi}-4\dot{\Phi}\dot{\phi}+\big(1+2\Phi\big)\frac{m^2}{\hbar^2}\phi=0,
\end{equation}
where $H=\Dot{a}(t)/a(t)$. At late times in an expanding universe, $H\ll m$ and it is sufficient to set $H=0$ and $a(t)=1$ in equation \ref{eq:linear}. This is a good approximation even at relatively high redshifts, including the epochs of early structure formation. Alternatively, if we consider a non-expanding universe, these equalities are true by definition. In either case, we can remove the second term in equation \ref{eq:linear} entirely, such that all remaining derivatives are second order in space or time.

In the non-relativistic limit, Equation \ref{eq:linear} can be treated using the WKB approximation\footnote{For a detailed explication of the WKB approximation in the non-relativistic limit, see \cite{Young2015}.}. This allows us to write an ansatz solution for the field $\phi$:
\begin{equation}\label{eq:ansatz}
    \phi=\frac{\hbar}{\sqrt{2}m}\big(\psi e^{-imt/\hbar}+\psi^* e^{imt/\hbar}\big),
\end{equation}
where $\psi$ is assumed to be slowly varying in the sense that $m\vert\psi\vert\gg\hbar\vert\dot{\psi}\vert$, $m\vert\dot{\psi}\vert\gg\hbar\vert\Ddot{\psi}\vert$, $m\vert\psi\vert\gg\hbar\vert\nabla{\psi}\vert$, and $m\vert\nabla{\psi}\vert\gg\hbar\vert\nabla^2{\psi}\vert$. Since $\Phi$ is sourced by $\psi$, we also have that $m\vert\Phi\vert\gg\hbar\vert\dot{\Phi}\vert$. Direct substitution of the ansatz solution into equation \ref{eq:linear}, discarding heavily suppressed terms, yields
\begin{equation}\label{eq:schrodinger}
    i\hbar\Dot{\psi}=-\frac{\hbar^2}{2m}\nabla^2\psi+m\Phi\psi.
\end{equation}
We have thus shown that $\psi$ satisfies the Schr{\"o}dinger equation in this limit, which is interpreted as the unique macroscopic wavefunction of a Bose-Einstein condensate. It follows that the particle density of the condensate is given by $\vert\psi\vert^2$, so its mass density is simply $m\vert\psi\vert^2$. The local gravitational potential thus satisfies the Poisson equation,
\begin{equation}\label{eq:poisson}
    \nabla^2\Phi=4\pi G m \vert\psi\vert^2,
\end{equation}
where $G$ is Newton's gravitational constant. The coupled equations \ref{eq:schrodinger} and \ref{eq:poisson} together form the nonlinear Schr{\"o}dinger-Poisson system which describes the dynamics of ULDM in the non-relativistic regime. It is this system of equations which \PyUltraLight is designed to solve.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth, trim={0 0 0 0},clip]{phase_comparison}
  \caption{Head-on collisions between solitons of mass 20, initial relative velocity 20, and initial separation 1.2 in code units. Time progresses from left to right across each row and is indicated in code units in each frame. The effective repulsive force generated by a $\pi$ phase shift can be seen in the lower panel.}
  \label{fig:repulsion}
\end{figure}

As demonstrated in \cite{Paredes2016}, the wavelike properties of ULDM give rise to effective forces which can dramatically affect the dynamics of core collisions. These effective forces arise as a result of interference phenomena, rather than because of any local interactions the ULDM model might  incorporate.  Figure \ref{fig:repulsion} shows the results of a head-on collision between two solitons, where in one instance the solitons have no initial phase difference, and in the other instance a phase difference of $\pi$ is applied in the initial conditions. In this simulation, solitons of mass 20 are initialised with relative velocity 20 and initial separation 1.2 (code units). The solitons are allowed to collide, and contours of the density profile along the plane of symmetry are displayed. In one case (top) there is no phase offset between the initial solitons, while in the second the phases differ by $\pi$. In the latter case, the phase shift creates an effective repulsive force between the two solitons. It can be seen in the second frame that as the solitons approach one another, the $\pi$ phase shift results in a slowing of the approach accompanied by a deformation of the density profile, acting so as to avoid contact between the solitons. Dissimilarly, in the case where there is no phase shift, the solitons readily collide and merge to form a single contiguous density profile prior to re-separating. Further discussion of this phenomenon and its possible observational consequences can be found in \cite{Paredes2016}.

@@

An extensive study of the tidal disruption of ULDM solitonic cores orbiting a central potential has recently been undertaken in \cite{Du2018} and we reproduce just one of their results here. When the Schr{\"o}dinger-Poisson system is recast as a system of hydrodynamical equations, the fluid velocity  at any point is the gradient of the smoothly-varying phase.  Since the curl of a gradient vanishes identically, the velocity field of the Schr{\"o}dinger-Poisson system must be irrotational, and $\nabla\times\vec{v}=0$. If a radially symmetric soliton is initialised in a circular orbit around a Newtonian potential, there will be initial transient behaviour as the spherical profile becomes elongated along the radial direction of the central potential. Meanwhile, the velocity field corresponding to the overall orbital motion of the soliton is superposed with the internal velocity field, combining so as to produce a net flow with vanishing curl.

The family of Riemann-S ellipsoids describe uniformly rotating bodies which satisfy the condition of irrotationality \cite{Chandrasekhar1965}, and it is therefore the characteristic internal velocity field of such an ellipsoid which we expect to arise during the simulation. It is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann-S ellipsoid when orbiting a host mass. We a reproduce this result using \PyUltraLight, with Figure \ref{fig:riemann} showing the internal velocity field of a solitonic satellite after one complete revolution around a host mass. The soliton has become elongated along the radial line connecting it to the host, indicating that it is tidally locked, while the internal velocity field qualitatively matches that of an irrotational Riemann-S ellipsoid within the tidal radius \cite{Daller2012}. It should be noted that the wider velocity field is not expected to be accurately predicted in a simulation of this kind. This is because the initial soliton density profile is defined only out to a given cutoff radius, beyond which the $\psi$ field value, and by extension its velocity field, is set identically to zero. Therefore, the velocity field far from the solitonic centre is not well-modelled in these simulations, and it may be expected that non-trivial self-interaction in the wider field would lead to unpredictable behaviour in these regions.

@@

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0.5cm 0 0,clip]{Total_Energy_Change_Disruption.eps}
  \caption{Demonstration of improving energy conservation with simulation resolution for a soliton undergoing significant tidal disruption in a Newtonian central potential.}
  \label{fig:energy_change_2}
\end{figure}

@@

Figures \ref{fig:combined_1} and \ref{fig:binary} demonstrate energy conservation for two scenarios. The first case  shows the evolution of the energy of a single soliton undergoing significant tidal disruption within a Newtonian central potential. For this simulation a soliton of mass 12 in code units was initialised at a radial distance of 3 code units from the centre of a Newtonian central potential generated by a central mass of 1000 code units. As the soliton is disrupted, the kinetic energy increases, while the gravitational energy due to the central potential decreases, as expected. Meanwhile, the gravitational potential energy from self-interactions gradually increases toward zero as the disruption continues and the ULDM is halo spread over a greater area. In this case the sum of the individual energy components is conserved to $10^{-4}\ \%$  at a resolution of $256^3$.

@@

Figure \ref{fig:energy_change_2} demonstrates the improvement in energy conservation with increasing grid resolution for the case of a single soliton becoming tidally disrupted in a Newtonian central potential. The simulation setup is the same as that used to generate Figure \ref{fig:combined_1}. Namely, a single soliton of mass 12 code units is initialised at a distance of 3 code units from a central mass, $M=1000$. The initial velocity of the soliton is $\sqrt{M/r}$ where $r$ is the radial distance of the soliton from the central mass. The duration of the simulation is 0.5 code units so that the soliton undergoes significant tidal disruption as demonstrated in Figure \ref{fig:combined_1}. While we see that energy is conserved at sub-percent level even for $64^3$ grid resolution, the qualitative behaviour of the mass density distribution in this case is not correct, so we conclude that this resolution is insufficient for convergence despite good energy conservation. This highlights the importance of a multifaceted approach to convergence testing. At $256^3$, energy is conserved to $10^{-4}\ \%$.

@@

The current implementation of \PyUltraLight is already a useful tool for simulating dynamical ULDM systems and exploring their dynamics However, there is much scope for  improvement. In particular, future releases may incorporate a variable timestep and more sophisticated physics, including explicit self-interactions in the axion sector or additional matter components.  \PyUltraLight is publicly available under a BSD license.

@@

\re{Check the bib options so we don't see all the Planck authors :-) }\ek{I've modified the bibliography style file so that only the first 3 authors will be  displayed if there are more than 3. I can't seem to find official JCAP guidance on this - Does it look OK to you in this format?}
